---
title: "The natural disasters that have most affected the United States"
author: "Brayam Pineda"
date: "30/4/2021"
output: html_document
---

## Abstract

The objetive of this analysis is to explore the NOAA Storm Database and answer some basic questions about severe weather events. In brief, the results show the "Tornado" was the most detrimental disaster in the list of events in the data. Hurricane/Typhoon are the second in the ranking.Furthermore, it was to plot two llolipop graphichs to show like the data is behaviored.

### Your data analysis must address the following questions:

#### 1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

#### 2. Across the United States, which types of events have the greatest economic consequences?

## Data Procesing

First one, import the libraries and determine a set.seed

```{r, message=FALSE}
library(readr)
library(dplyr)

set.seed(1234)
```

Secod, call our data

```{r, cache=TRUE, message=FALSE, warning=FALSE}
path = "repdata_data_StormData.csv.bz2"
data <- read_csv(path)
head(data)
```

### Now, for the first question

It is necesary remake the data base to respond the question, in this case we need to count the number of the injuries by type of the event.


```{r}
evtype <- data %>% group_by(EVTYPE) %>% summarise(INJURIES = sum(INJURIES))
evtype <- evtype[order(evtype$INJURIES, decreasing = TRUE),]  
head(evtype)
```

Now, we can to plot this information to do clearer the information, by avoiding the classic histogram it is posible plot a llolipop graph.

```{r}
library(ggplot2)
head(evtype, 20) %>%
  arrange(INJURIES) %>%    # First sort by val. This sort the dataframe but NOT the factor levels
  mutate(name = factor(EVTYPE, levels = EVTYPE)) %>%   # This trick update the factor levels
  ggplot( aes(x = name, y = INJURIES)) +
  geom_segment( aes(xend = name, yend=0)) +
  geom_point( size=4, color="orange") +
  coord_flip() +
  theme_bw() +
  xlab("")
```

In both graphs it is possible to observe that the type of event that causes the most injueries is the tornado followed by far by TSTM wind.

### The second question is 

Across the United States, which types of events have the greatest economic consequences?

To answer is necesary to recode some variable in order to be able to be understandable by the program.

```{r}
data$PROPDMGEXP[data$PROPDMGEXP == ""] <- NA
data$PROPDMGEXP <- gsub("1", "10", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("NA", "1", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("\\-", "1", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("\\?", "1", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("\\+", "1", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("^0$", "1", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("h|H", "100", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("K", "1000", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("m|M", "1000000", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("B", "1000000000", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("2", "100", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("3", "1000", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("4", "10000", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("5", "100000", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("6", "1000000", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("7", "10000000", data$PROPDMGEXP)
data$PROPDMGEXP <- gsub("8", "100000000", data$PROPDMGEXP)


data$PROPDMGEXP <- as.numeric(data$PROPDMGEXP)
```

After that is posible to remake the data to get clarer information.

```{r}
propdmgexp <- data %>% group_by(EVTYPE) %>% summarise("PROPERTY DAMAGED" = sum(PROPDMGEXP, na.rm = TRUE))
propdmgexp <- propdmgexp[order(propdmgexp$`PROPERTY DAMAGED`, decreasing = TRUE), ]
head(propdmgexp)
```

Now the information is ready to the visualization

```{r}
head(propdmgexp[propdmgexp$EVTYPE != "?", ], 20) %>%
  arrange(`PROPERTY DAMAGED`) %>%    # First sort by val. This sort the dataframe but NOT the factor levels
  mutate(name = factor(EVTYPE, levels = EVTYPE)) %>%   # This trick update the factor levels
  ggplot( aes(x = name, y = `PROPERTY DAMAGED`)) +
  geom_segment( aes(xend = name, yend=0)) +
  geom_point( size=4, color="orange") +
  coord_flip() +
  theme_bw() +
  xlab("")
```

## Results

As was possible to observe from the graphics the Tornado was the more harmful event which to affected to United States in the past years, however hurricane or typhoon were the disaster which more affected to economic. Also Tornado was included in the last actegory where it get the second place. In brief Tornado is the most detrimental type of event in the information.